mathml: Translate R Expressions to MathML and LaTeX

### 2023-12-21 rj/mathml.Rmd
IA: draft changes

---
title: "`mathml`: Translate R Expressions to MathML and LaTeX"
date: "2023-11-20"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the R\ Markdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package enables use of the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks enable extension of the package by
  mapping specific variables or functions to new MathML and LaTeX entities. 
  Throughout the paper, examples are given for the functions of the package, and
  a case study illustrates its use in a scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
---

@@

To avoid name clashes with package `stats`, `roof()` is used to put a hat
on a symbol (see next section for further decorations). Note that an 
R\ function `roof()` does not exist in base R, it is provided by the package
for convenience and points to the identity function.

@@

The package offers some support for different fonts as well as accents and
boxes etc. Internally, these decorations are implemented as identity functions,
so that they can be introduced into R expressions without side-effects.

@@

The example is a reminder that it is not possible to define the precedence of
custom operators in R, and that expressions with such operators are evaluated
strictly from left to right. Again, the problem can be worked around by the
functional form of the operator or a curly brace to hide the parenthesis, and,
at the same time, enforce the correct operator precedence.

@@

For self-written functions, the matter is somewhat more complicated. For a function
such as `g <- function(...) ...`, the name _g_ is not transparent to R, because
only the function body is represented. We can still display functions in the
form `head(x) = body` if we embed the object to be shown into a
call `"<-"(head, body)`.

@@

Note that in the example above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. While the R\ function `integrate()` can handle both symbols and
functions, \pkg{mathml} needs the symbol because it is unable to determine the
function name of custom functions.

@@

Note that the seemingly more convenient `term <- quote(A + B)` yields \(A + B\)
in the output---instead of the desired matrix representation. This behavior is
expected because quotation of R calls also quote the components of the
call (here, _A_ and _B_).

@@

The R function `omit_right(a + b)` uses non-standard evaluation
techniques [e.g., @Wickham2019] to return only the left part an operation,
and cancels the right part. This may not always be desired, for example, when
illustrating how to fix the mistake.

@@

One would typically use chunk option `echo=FALSE` to suppress the R code.

@@

One limitation of the package is the lack of a convenient way to insert line
breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
For example, in its current stage, the LaTeX package breqn [@breqn] is mostly
a proof of concept. Moreover, \pkg{mathml} only works in one direction,
that is, it is not possible to translate from LaTeX or HTML back to R 
[see @latex2r, for an example].

@@

For convenience, the translation of the R expressions is achieved through a
Prolog interpreter provided by another R\ package \CRANpkg{rolog} [@rolog]. If a
version of SWI-Prolog [@swipl] is found on the system, \CRANpkg{rolog} connects
to it. Alternatively, the SWI-Prolog runtime libraries can be conveniently
accessed by installing the R\ package \CRANpkg{rswipl} [@rswipl]. Prolog is a
classical logic programming language with many applications in expert systems,
computer linguistics and symbolic artificial intelligence. The strength of
Prolog lies in its concise representation of facts and rules for knowledge and
grammar, as well as its efficient built-in search engine for closed world
domains. Whereas Prolog is weak in statistical computation, but strong in
symbolic manipulation, the converse may be said for the R\ language. \pkg{rolog}
bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two Prolog functions allow ring
back and evaluation of terms in R.

The proper term for a Prolog "function" is predicate, and it is typically
written with name and arity (i.e., number of arguments), separated by a forward
slash. Thus, at the Prolog end, the predicate `math/2` translates the
representation of the R call `pbinom(K, N, Pi)` into a more general 
representation of an R function `fn/2` with the name `P_Bi`, one 
argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.

@@

Several ways exist for translating new R\ terms to their mathematical
representation. We have already seen above how to use "hooks" to translate long
variable names from R to compact mathematical signs, as well as functions such
as cumulative probabilities \(P(X \le k)\) to different representations
like \(\sum_{i=0}^k P(X = i)\). Obviously, the hooks require that there already
exists a rule to translate the target representation into MathML and MathJax.

@@

@@

Base R does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
the present package does not support the respective representation. To obtain a
cube root, a programmer would typically type `x^(1/3)` or better `x^{1/3}` (see
the practice section why the curly brace is preferred in an exponent), resulting
in \(x^{1/3}\) which may still not match everyone's taste. Here we describe the
steps needed to represent the \(n\)-th root as \(\sqrt[n]x\).

We assume that `nthroot(x, n)` is available in the current
namespace [manually defined, or from R package \CRANpkg{pracma}, @pracma], so that
the names of the arguments and their order are accessible to `canonical()` if
needed. As we can see below, \pkg{mathml} uses a default
representation `name(arguments)` for such unknown functions.

>>>>

---
title: "mathml: Translate R Expressions to MathML and LaTeX"
date: "2023-11-20"
abstract: >
  This R\ package translates R\ objects to suitable elements in MathML or LaTeX,
  thereby allowing for a pretty mathematical representation of R\ objects and
  functions in data analyses, scientific reports and interactive web content. In
  the R\ Markdown document rendering language, R\ code and mathematical content
  already exist side-by-side. The present package enables use of the same
  R\ objects for both data analysis and typesetting in documents or web content.
  This tightens the link between the statistical analysis and its verbal
  description or symbolic representation, which is another step towards
  reproducible science. User-defined hooks enable extension of the package by
  mapping specific variables or functions to new MathML and LaTeX entities. 
  Throughout the paper, examples are given for the functions of the package, and
  a case study illustrates its use in a scientific report.
draft: true
author:  
  - name: Matthias Gondan
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/gondan-rochon/index.html.en
    orcid: 0000-0001-9974-0057
    email:  Matthias.Gondan-Rochon@uibk.ac.at
  - name: Irene Alfarone
    url: https://www.uibk.ac.at/psychologie/mitarbeiter/alfarone/index.html.en
    email: Irene.Alfarone@uibk.ac.at
    orcid: 0000-0002-8409-8900
    affiliation: Universität Innsbruck
    address:
    - Department of Psychology
    - Innsbruck, Austria
type: package
output: 
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
bibliography: bibliography.bib
preamble:
- \usepackage{cancel}
- \usepackage{amsmath}
---

@@

To avoid name clashes with package `stats`, `roof()` is used to put a hat
on a symbol (see next section for further decorations). Note that an 
R\ function `roof()` does not exist in base R\, it is provided by the package
for convenience and points to the identity function.

@@

The package offers some support for different fonts as well as accents and
boxes etc. Internally, these decorations are implemented as identity functions,
so that they can be introduced into R\ expressions without side-effects.

@@

The example is a reminder that it is not possible to define the precedence of
custom operators in R\, and that expressions with such operators are evaluated
strictly from left to right. Again, the problem can be worked around by the
functional form of the operator or a curly brace to hide the parenthesis, and,
at the same time, enforce the correct operator precedence.

@@

For self-written functions, the matter is somewhat more complicated. For instance, 
if we consider a function such as `g <- function(...) ...`, the name _g_ is not 
transparent to R, because only the function body is represented. 
We can still display functions in the form `head(x) = body` if we embed the object to be shown into a call `"<-"(head, body)`.

@@

Above, the quoted term is an abbreviation
for `call("integrate", quote(sin), ...)`, with `sin` being an R\ symbol, not a
function. While the R\ function `integrate()` can handle both symbols and
functions, \pkg{mathml} needs the symbol because it is unable to determine the
function name of custom functions.

@@

Note that the seemingly more convenient `term <- quote(A + B)` yields \(A + B\)
in the output---instead of the desired matrix representation. This behavior is
expected because quotation of R\ calls also quote the components of the
call (here, _A_ and _B_).

@@

The R\ function `omit_right(a + b)` uses non-standard evaluation
techniques [e.g., @Wickham2019] to return only the left part an operation,
and cancels the right part. This may not always be desired, for example, when
illustrating how to fix the mistake.

@@

One would typically use chunk option `echo=FALSE` to suppress the R\ code.

@@

One limitation of the package is the lack of a convenient way to insert line
breaks. This is mostly due to lacking support by MathML and LaTeX renderers. 
For example, in its current stage, the LaTeX package breqn [@breqn] is mostly
a proof of concept. Moreover, \pkg{mathml} only works in one direction,
that is, it is not possible to translate from LaTeX or HTML back to R\ 
[see @latex2r, for an example].

@@

For convenience, the translation of the R\ expressions is achieved through a
Prolog interpreter provided by another R\ package \CRANpkg{rolog} [@rolog]. If a
version of SWI-Prolog [@swipl] is found on the system, \CRANpkg{rolog} connects
to it. Alternatively, the SWI-Prolog runtime libraries can be conveniently
accessed by installing the R\ package \CRANpkg{rswipl} [@rswipl]. Prolog is a
classical logic programming language with many applications in expert systems,
computer linguistics and symbolic artificial intelligence. The strength of
Prolog lies in its concise representation of facts and rules for knowledge and
grammar, as well as its efficient built-in search engine for closed world
domains. Whereas Prolog is weak in statistical computation, but strong in
symbolic manipulation, the converse may be said for the R\ language. \pkg{rolog}
bridges this gap by providing an interface to a SWI-Prolog
distribution [@swipl] in R. The communication between the two systems is mainly
in the form of queries from R\ to Prolog, but two Prolog functions allow ring
back and evaluation of terms in R.

The proper term for a Prolog "function" is predicate, and it is typically
written with name and arity (i.e., number of arguments), separated by a forward
slash. Thus, at the Prolog end, the predicate `math/2` translates the
representation of the R\ call `pbinom(K, N, Pi)` into a more general 
representation of an R\ function `fn/2` with the name `P_Bi`, one 
argument `X =< K`, and the two parameters `N` and `Pi`, as shown below.

@@

Several ways exist for translating new R\ terms to their mathematical
representation. We have already seen above how to use "hooks" to translate long
variable names from R\ to compact mathematical signs, as well as functions such
as cumulative probabilities \(P(X \le k)\) to different representations
like \(\sum_{i=0}^k P(X = i)\). Obviously, the hooks require that there already
exists a rule to translate the target representation into MathML and MathJax.

@@

See Section 3.11 for replacing lengthy R\ labels by short mathematical symbols.

@@

Base R\ does not provide a function like `cuberoot(x)` or `nthroot(x, n)`, and
the present package does not support the respective representation. To obtain a
cube root, a programmer would typically type `x^(1/3)` or better `x^{1/3}` (see
the practice section why the curly brace is preferred in an exponent), resulting
in \(x^{1/3}\) which may still not match everyone's taste. Here we describe the
steps needed to represent the \(n\)-th root as \(\sqrt[n]x\).

We assume that `nthroot(x, n)` is available in the current
namespace [manually defined, or from R\ package \CRANpkg{pracma}, @pracma], so that
the names of the arguments and their order are accessible to `canonical()` if
needed. As we can see below, \pkg{mathml} uses a default
representation `name(arguments)` for such unknown functions.
